``Interestingness'' in Mathematical Theory Exploration

### 2019-11-19 PhDThesis/related_formalstats.tex
Edits

An important property of a logical formula (in general, and in particular with
regards to interestingness) is its truth value. Although we may be able to
determine some truth values exactly, e.g. via deduction, it may only be possible
(or more efficient) to \emph{approximate} truth values. One straightforward
approximation is to adopt \emph{probabilities}, where we assign probability $1$
to formulae which are known to be true (e.g. axioms), $0$ to formulae known to
be false, and intermediate values to those which are not known.

This can be useful for ``knowledge bases'' of non-logical axioms (e.g. the
observed colours of Hempel's ravens~\cite{10.1093/mind/LIV.214.97} \iffalse
TODO: Alison: Maybe a bit of explanation here \fi), and allows the probability of
universal statements to be inferred from non-exhaustive
cases~\cite{Hutter.Lloyd.Ng.ea:2013}. However, inferring consistent
probabilities requires \emph{logical omniscience}: knowledge of all the logical
consequences of the chosen axioms. This is unfortunate, since it is precisely
what we are trying to avoid by approximating with probabilities.

Avoiding this requirement for logical omniscience has given rise to the field of
\emph{bounded rationality}, where reasoning (deduction, induction, etc.) is
subject to some computational constraints. For example, the ``logical inductor''
of~\cite{garrabrant2016logical} avoids assigning particular probabilities to
formal statements, and instead produces a \emph{sequence} of successive
approximate probabilities, each computable with finite (though
doubly-exponential!) resources. These approximations are only guaranteed to be
consistent in the limit.

@@

>>>>

An important property of a logical formula (in general, and in particular with
regards to interestingness) is its truth value. Although we may sometimes be
able to determine truth values exactly, e.g. via deduction, for many formulae it
may only be possible (or practical) to \emph{approximate} their truth value. One
straightforward approximation is to adopt \emph{probabilities}, where we assign
probability $1$ to formulae which are known to be true (e.g. axioms), $0$ to
formulae known to be false, and intermediate values to those which are not
known~\cite{Hutter.Lloyd.Ng.ea:2013}. For example, the truth value of universal
statements (e.g. ``all ravens are black'') cannot be deduced from non-exhaustive
examples (e.g. real-world observations of raven colours), but various induction
methods may infer an (approximate) truth value~\cite{10.1093/mind/LIV.214.97}.

One difficulty with approximate truth values is maintaining consistency without
requiring \emph{logical omniscience} (knowledge of all the logical consequences
of the chosen axioms). Avoiding logical omniscience has given rise to the field
of \emph{bounded rationality}, where reasoning (deduction, induction, etc.) is
subject to some computational constraints. For example, the ``logical inductor''
of~\cite{garrabrant2016logical} avoids assigning particular probabilities to
formal statements, and instead produces a \emph{sequence} of successive
approximate probabilities, each computable with finite resources (albeit very
inefficiently), and guaranteed to be \emph{eventually consistent}.

@@

To avoid such trivially-easy and infeasibly-difficult statements (which we may
consider uninteresting) we can instead consider the \emph{variance}
(disagreement) between \emph{multiple} approximate truth values (similar to the
problem solvers of~\cite{fernando2013design1, fernando2013design2}). Those
statements which are assigned similar truth values by a variety of methods may
be considered less interesting, as they are either easy enough to infer by all
of our methods, or too hard for any of them (the stability of a single method's
prediction over time would serve a similar role). One benefit of this
application is that the approximation methods do not have to be particularly
capable; in fact choosing ``superhuman'' algorithms would be \emph{less} useful
in this context, since they would agree on the triviality of many statements we
might consider interesting.
